Data Science Applied to Ag - Final Project - ML

### 2025-04-25 code/Draft_script.qmd
d

@@

@@

```{r clean_and_summarize_trait, message=F, warning=F}
library(stringr)

trait_avg_df <- trait_df %>%
  select(-block) %>%
  # 1) remove all lowercase letters from site
  # 2) keep only text before any dash
  # 3) keep only text before any underscore
  mutate(
    site = str_remove_all(site, "[a-z]"),
    site = str_replace(site, "-.*$", ""),
    site = str_replace(site, "_.*$", "")
  ) %>%
  group_by(year, site, hybrid) %>%
  summarize(
    yield_mg_ha    = mean(yield_mg_ha, na.rm = TRUE),
    grain_moisture = mean(grain_moisture, na.rm = TRUE),
    date_planted   = first(date_planted),
    date_harvested = first(date_harvested),
    .groups = "drop"
  ) %>%
  mutate(
    adjusted_yield = adjust_yield(yield_mg_ha, grain_moisture),
    planting_date  = as.Date(date_planted,   "%m/%j/%y"),
    harvest_date   = as.Date(date_harvested, "%m/%j/%y"),
    plant_doy      = yday(planting_date),
    harvest_doy    = yday(harvest_date),
    season_len     = harvest_doy - plant_doy
  ) %>%
  select(-yield_mg_ha, -grain_moisture, -date_planted, -date_harvested) %>%
  ungroup()

@@

write_csv(trait_avg_df,
          "../data/training_trait_clean.csv")
```

@@

soil_clean <- soil_df %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  )
```

@@

write_csv(soil_clean,
          "../data/training_soil_clean.csv")
```

@@

meta_clean <- meta_df %>%
  rename(
    lon = longitude,
    lat = latitude
  ) %>%
  select(-previous_crop) %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  ) %>%
  distinct(year, site, .keep_all = TRUE) %>%
  ungroup()

@@

@@

write_csv(meta_clean,
          "../data/training_meta_clean.csv")
```

@@

write_csv(merged_clean,
          "../data/training_merged_clean.csv")
```

@@

write_csv(daymet_all_unnest,
          "../data/merged_fieldweatherdata.csv"
          )

@@

fieldweather <- read_csv("../data/merged_fieldweatherdata.csv")

fieldweather

@@

write_csv(fe_month_sum_wide,
          # "../data/weather_monthsum.csv")

@@

>>>>

# Start of data wrangling

@@

## EDA

summary(trait_df)

```
```{r}

summary(meta_df)

summary(soil_df)

# Data wrangling on "trait_df"

The code below creates a function to adjust the yield_mg_ha for 15.5% grain moisture.

@@

The code below conducts data wrangling on "trait_df"

```{r clean_and_summarize_trait, message=F, warning=F}

trait_clean <- trait_df %>%
  select(-block) %>%
  mutate(
    site = str_remove_all(site, "[a-z]"), # removing all lowercase letters from site
    site = str_replace(site, "-.*$", ""), # keeping only text before any dash
    site = str_replace(site, "_.*$", "") # keeping only text before any underscore
  ) %>%
  group_by(year, site, hybrid) %>%
  summarize(
    yield_mg_ha    = mean(yield_mg_ha, na.rm = TRUE),
    grain_moisture = mean(grain_moisture, na.rm = TRUE),
    date_planted   = first(date_planted),
    date_harvested = first(date_harvested),
    .groups = "drop"
  ) %>%
  mutate(
    adjusted_yield = adjust_yield(yield_mg_ha, grain_moisture),
    planting_date  = as.Date(date_planted,   "%m/%j/%y"),
    harvest_date   = as.Date(date_harvested, "%m/%j/%y"),
    plant_doy      = yday(planting_date),
    harvest_doy    = yday(harvest_date),
  ) %>%
  select(-yield_mg_ha, -grain_moisture, -date_planted, -date_harvested, -planting_date, -harvest_date) %>%
  ungroup()

@@

#write_csv(trait_avg_df,
          #"../data/training_trait_clean.csv")
```

# Data wrangling on "soil_df"

@@

soil_clean <- soil_df %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  ) %>%
  ungroup()

summary(soil_clean)
```

@@

#write_csv(soil_clean,
          #"../data/training_soil_clean.csv")
```

@@

meta_clean <- meta_df %>%
  rename(
    lon = longitude,
    lat = latitude
  ) %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  ) %>%
  distinct(year, site, .keep_all = TRUE) %>%
  ungroup()

@@

The code below will create categories for previous crops.

```{r}
library(stringr)

meta_clean <- meta_clean %>%
  mutate(
    # 1) lowercase & trim
    prev_crop_raw = str_to_lower(str_squish(previous_crop)),
    # 2) recode into your desired levels
    prev_crop = case_when(
      prev_crop_raw == "cotton"                                   ~ "cotton",
      prev_crop_raw == "peanut"                                   ~ "peanut",
      prev_crop_raw %in% c("soybean", "soybeans")                 ~ "soybean",
      prev_crop_raw == "sorghum"                                  ~ "sorghum",
      prev_crop_raw %in% c("wheat", "winter wheat", "2019/20 wheat") ~ "wheat",
      TRUE                                                         ~ "others"
    ) %>%
    # 3) make it a factor with the exact ordering you want
    factor(levels = c("cotton", "peanut", "soybean", "sorghum", "wheat", "others"))
  ) %>%
  select(-previous_crop, -prev_crop_raw) %>%
  clean_names()

meta_clean

```{r}
summary(meta_clean)
```

@@

#write_csv(meta_clean,
          #"../data/training_meta_clean.csv")
```

@@

#write_csv(merged_clean,
          #"../data/training_merged_clean.csv")
```

@@

#write_csv(daymet_all_unnest,
          #"../data/merged_fieldweatherdata.csv"
          )

@@

#fieldweather <- read_csv("../data/merged_fieldweatherdata.csv")

#fieldweather

@@

#write_csv(fe_month_sum_wide,
          # "../data/weather_monthsum.csv")

@@

# Feature engineering "planting stages" and "harvest stages":

The code chunk below is to feature engineer planting and harvest stages based on the following cut-offs for planting day of the year (plant_doy) and harvest day of the year (harvest_doy):

Planting stages:
 - early: plant_doy ≤ 120
 - normal: plant_doy = 121–150
 - late: plant_doy > 150

Harvest stages:
 - early: harvest_doy ≤ 275
 - normal: harvest_doy = 276–305
 - late: harvest_doy > 305

trait_avg_df <- trait_avg_df %>%
mutate(
    planting_stages = case_when(
      plant_doy <= 120                     ~ "early",
      plant_doy >= 121 & plant_doy <= 150  ~ "normal",
      plant_doy > 150                      ~ "late"
    ) %>% 
      fct_relevel("early","normal","late"),

    harvest_stages = case_when(
      harvest_doy <= 275                    ~ "early",
      harvest_doy >= 276 & harvest_doy <= 305 ~ "normal",
      harvest_doy > 305                     ~ "late"
    ) %>% 
      fct_relevel("early","normal","late")
  ) %>%
  select( -plant_doy, -harvest_doy) %>%
  ungroup()

View(trait_avg_df)
